# Visualization

For visualization, we're going to use `ggplot2` - a powerful set of commands for plot generation. Let's make sure we've got our data from the last chapter active in our current session:

```{r}
library(tidyverse)
algae_chemistry <- read_csv("https://thebustalab.github.io/R_For_Chemists/sample_data/algae_data.csv")
algae_chemistry_small <- filter(algae_chemistry, chemical_species == "FAs")
algae_chemistry_small
```

Great! Looks like we're ready to go.

## Setting up a ggplot

There are three steps to setting up a ggplot:

### Step 1: Define the data you want to use.

We do this using the ggplot function's data argument. When we run that line, it just shows a grey plot space. Why is this? It's because all we've done is told ggplot that (i) we want to make a plot and (ii) what data should be used. We haven't explained how to represent features of the data using ink.

```{r}
ggplot(data = algae_chemistry_small)
```

### Step 2: Define how your variables map onto the axes.

This is called aesthetic mapping and is done with the `aes()` function. `aes()` should be placed inside the `ggplot` command. Now when we run it, we get our axes!

```{r}
ggplot(data = algae_chemistry_small, aes(x = algae_strain, y = abundance))
```

### Step 3: Use geometric shapes to represent other variables in your data.

Map your variables onto the geometric features of the shapes. To define which shape should be used, use a `geom_*` command. Some options are, for example, `geom_point()`, `geom_boxplot()`, and `geom_violin()`. These functions should be added to your plot using the `+` sign. We can use a new line to keep the code from getting too wide, just make sure the `+` sign is at the end fo the top line. Again, use `aes()` to map your variables onto the geometric features of the shapes. Let's try it:

```{r}
ggplot(data = algae_chemistry_small, aes(x = algae_strain, y = abundance)) + 
  geom_point(aes(color = harvesting_regime))
```

## Geoms

### Modifying geoms

In the last plot in the previous section, the points were a bit small, how could we fix that? We can modify the features of the shapes by adding additional arguments to the `geom_*()` functions. To change the size of the points created by the `geom_point()` function, this means that we need to add the `size = ` argument. Here's an example:

```{r}
ggplot(data = algae_chemistry_small, aes(x = algae_strain, y = abundance)) + 
  geom_point(aes(color = harvesting_regime), size = 5)
```

One powerful aspect of `ggplot` is the ability to quickly change mappings to see if alternative plots are more effective at bringing out the trends in the data. For example, we could modify the plot above by switching how harvesting_regime is mapped:

```{r}
ggplot(data = algae_chemistry_small, aes(x = algae_strain, y = abundance)) +
	geom_point(aes(size = harvesting_regime), color = "black")
```

** Important note: Inside the `aes()` function, map aesthetics (the features of the geom's shape) to a *variable*. Outside the `aes()` function, map aesthetics to *constants*. You can see this in the above two plots - in the first one, color is inside `aes()` and mapped to the variable called harvesting_regime, while size is outside the `aes()` call and is set to the constant 5. In the second plot, the situation is reversed, with size being inside the `aes()` function and mapped to the variable harvesting_regime, while color is outside the `aes()` call and is mapped to the constant "black".

### Using multiple geoms

We can also stack geoms on top of one another by using multiple `+` signs. We also don't have to assign the same mappings to each geom.

```{r}
ggplot(data = algae_chemistry_small, aes(x = algae_strain, y = abundance)) + 
  geom_violin() +
  geom_point(aes(color = harvesting_regime), size = 5)
```

As you can probably guess right now, there are lots of mappings that can be done, and lots of different ways to look at the same data!

```{r}
ggplot(data = algae_chemistry_small, aes(x = algae_strain, y = abundance)) +
	geom_violin(aes(fill = algae_strain)) +
  geom_point(aes(color = harvesting_regime, size = replicate))
```