# specialized plots and maps {-}

## {-}

This appendix collects plot types that are useful but specialized. None of them is part of the main
sequence of the book — each is a self-contained recipe you can come back to when your data happens to
have the shape it fits. Three of them (3D scatter, Venn, ternary) are unusual chart forms; the fourth,
mapping, is a larger topic in its own right.

*These sections were the tail of the former "data visualization III" chapter, which was dissolved on
2026-08-27: its network material became chapter 7, its distribution material moved into the
comparing-means chapter, and what remained is here.*

## specialized plots {-}

### 3D scatter plots {-}

`phylochemistry` contains a function to help you make somewhat decent 3D scatter plots. Let's look at an example (see below). For this, we use the function `points3D`. We give it a `data` argument that gives it vectors of data that should be on the x, y, and z axes, along with a vector that uniquely identifies each observation. We also tell it the angle of the z axis that we want, the integer to which ticks should be rounded, and the tick intervals. The function returns data that we can pass to ggplot to make a 3D plot.

```{r}
pivot_wider(hawaii_aquifers, names_from = "analyte", values_from = "abundance") %>%
  mutate(sample_unique_ID = paste0(aquifer_code, "_", well_name)) -> aquifers

output <- points3D(
  data = data.frame(
    x = aquifers$SiO2,
    y = aquifers$Cl,
    z = aquifers$Mg,
    sample_unique_ID = aquifers$sample_unique_ID
  ),
  angle = pi/2.4,
  tick_round = 10,
  x_tick_interval = 10,
  y_tick_interval = 20,
  z_tick_interval = 20
)

str(output)
```

The output from points3D contains a grid, axes, and ticks, which should all be plotted using geom_segment. It also contains points that should be plotted with geom_point, and point segments that should be plotted with geom_segment. We can take the output from points3D and join it with the original data, which will occur according to our sample_unique_ID column. Then, we can also plot point metadata:

```{r}
output$points <- left_join(output$points, aquifers)
  
ggplot() +
  geom_segment(
    data = output$grid, aes(x = x, xend = xend, y = y, yend = yend),
    color = "grey80"
  ) +
  geom_segment(data = output$axes, aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_segment(data = output$ticks, aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_text(
    data = output$labels, aes(x = x, y = y, label = label),
    hjust = 0.5
  ) +
  geom_segment(
    data = output$point_segments,
    aes(x = x, xend = xend, y = y, yend = yend),
    linetype = "dotted", color = "black"
  ) +
  geom_point(
    data = output$points, aes(x = x, y = y, fill = aquifer_code),
    size = 3, shape = 21
  ) +
  theme_void() +
  scale_fill_manual(values = discrete_palette)
```

### venn diagrams {-}

```{r}
df <- data.frame(
  plant1 = sample(c(TRUE, FALSE), 24, replace = TRUE),
  plant2 = sample(c(TRUE, FALSE), 24, replace = TRUE),
  plant3 = sample(c(TRUE, FALSE), 24, replace = TRUE),
  attribute_name = sample(letters, 24, replace = FALSE)
)

vennAnalysis(df[,1:3]) %>%
  ggplot() +
    geom_circle(
      aes(x0 = x, y0 = y, r = r, fill = category),
      alpha = 0.4
    ) +
  scale_fill_brewer(palette = "Set1") +
  theme_void()
```


### ternary plots {-}

```{r}
library(ggplot2)
library(ggtern)
alaska_lake_data %>%
  pivot_wider(names_from = "element", values_from = "mg_per_L") %>%
  ggtern(aes(
    x = Ca,
    y = S,
    z = Na,
    color = park,
    size = pH
    )) +
  geom_point() 
```



## map data {-}

Geographic data is really its own subject, and what follows is an introduction rather than a treatment
of it. The point to take away is that a map is a plot like any other — the coordinates happen to be
longitude and latitude, and the one genuinely new idea is the **projection**, which is the decision
about how to flatten a curved surface onto a page. Every projection distorts something; choosing one is
a decision with consequences, not a default.

### plotting boundaries {-}

There is a simple way to plot maps with ggplot. The map data comes with `ggplot2`! Let's have a look. See below some of the data sets included. Options included with ggplot are: `world`, `world2`, `usa`, `state` (US), `county` (US), `nz`, `italy`, and `france`. `geom_polygon()` is useful for plotting these, at (at least to me) seems more intuitive than `geom_map()`.

```{r}
head(map_data("world"))
```
```{r}
head(map_data("state"))
```

```{r}
head(map_data("county"))
```

```{r}
head(map_data("france"))
```

Cool! We can see that lat, lon, group, order, region, and subregion are included. That makes plotting easy. Note that `coord_map()` can help preserve aspect ratios:

```{r, fig.height = 8, fig.width = 10}
ggplot(map_data("world")) +
  geom_point(aes(x = long, y = lat, color = group), size = 0.5) +
  theme_void() +
  coord_map()
```

Note that we can use `coord_map()` to do some pretty cool things!

```{r, fig.height = 8, fig.width = 10}
ggplot(map_data("world")) +
  geom_point(aes(x = long, y = lat, color = group), size = 0.5) +
  theme_void() +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45)
```

We can use filtering to produce maps of specific regions.

```{r, fig.height = 8, fig.width = 10}
ggplot() +
  geom_polygon(
    data = filter(map_data("county"), region == "minnesota"),
    aes(x = long, y = lat, group = subregion, fill = subregion),
    color = "black"
  ) +
  theme_void() +
  coord_map()
```

### maps with plots {-}

Please note that the Great Lakes are in map_data()!

```{r}
filter(map_data("lakes"), region == "Great Lakes", subregion == "Superior") %>%
    ggplot() +
      geom_path(aes(x = long, y = lat)) +
      coord_map() +
      theme_minimal()
```

We can clean up the map by making different groups for geom_path() whenever two consecutive points are far apart:

```{r}
# Step 1: Filter and prepare your data for Lake Superior (though yes, that includes Michigan and Huron)
lake_superior <- map_data("lakes") %>%
  filter(region == "Great Lakes", subregion == "Superior") %>%
  arrange(order)

# Step 2: Calculate distances between consecutive points
lake_superior <- lake_superior %>%
  mutate(lag_long = lag(long),
         lag_lat = lag(lat),
         dist_to_prev = geosphere::distHaversine(cbind(long, lat), cbind(lag_long, lag_lat))) # distHaversine calculates distances

# Step 3: Define a threshold (e.g., 50 km) and create the "distance_group"
threshold <- 50000  # 50 km
lake_superior <- lake_superior %>%
  mutate(distance_group = cumsum(ifelse(dist_to_prev > threshold | is.na(dist_to_prev), 1, 0)))

# Step 4: Plot the map with `distance_group`
ggplot(lake_superior, aes(x = long, y = lat, group = distance_group)) +
  geom_path() +
  coord_map() +
  theme_minimal()
```

Now we could add some data. The next few examples use a dataset of per- and polyfluoroalkyl substance (PFAS) measurements from sites around Lake Superior. **Note: these are unpublished data from ongoing lab research, included here purely to illustrate the plotting techniques. The file is not distributed with the course, so the code below is shown for reference and will not run on your machine — focus on the mapping and layering approach rather than reproducing the figure.** We could do something simple like plot total abundances as the size of a point:

```{r, fig.width = 8, fig.height = 4}
lake_superior_PFAS <- readMonolist("/Users/bust0037/Documents/Tools/websites/private_data/pfas_data_private.csv")
lake_superior_PFAS %>%
  group_by(site, lon, lat) %>%
  summarize(total = sum(abundance)) -> lake_superior_PFAS_summarized

ggplot() +
  geom_path(
    data = filter(lake_superior, lat > 46, long < -84),
    aes(x = long, y = lat, group = distance_group)
  ) +
  geom_point(
    data =  lake_superior_PFAS_summarized,
    aes(x = lon, y = lat, size = total),
    color = "black"
  ) +
  coord_map() +
  theme_cowplot()
```

Or we could do something more sophisticated like add pie charts at each point:


```{r, fig.width = 8, fig.height = 4}
lake_superior_PFAS <- readMonolist("/Users/bust0037/Documents/Tools/websites/private_data/pfas_data_private.csv")

grouped_by_site <- filter(lake_superior_PFAS, component == "PFBA")
site_less_than_90lon <- filter(grouped_by_site, lon <= -90)
site_more_than_90lon <- filter(grouped_by_site, lon >= -90)

unique_sites <- unique(lake_superior_PFAS$site)
dataframe_of_pies <- list()
for (i in 1:length(unique_sites)) { #i=1
  this_site <- filter(lake_superior_PFAS, site == unique_sites[i])
  this_site %>%
    ggplot()+
    geom_col(aes(x = 1, y = abundance, fill = class_name), color = "black") +
    coord_polar(theta = "y") +
    theme_void() +
    scale_fill_brewer(palette = "Set1", guide = "none") -> this_sites_pie
  dataframe_of_pies[[i]] <- tibble(x = this_site$lon[1], y = this_site$lat[1], plot = list(this_sites_pie))
}
dataframe_of_pies <- do.call(rbind, dataframe_of_pies)

ggplot() +
  geom_path(
    data = filter(lake_superior, lat > 46, long < -84),
    aes(x = long, y = lat, group = distance_group)
  ) +
  geom_point(
    data =  lake_superior_PFAS,
    aes(x = lon, y = lat),
    color = "black"
  ) +
  geom_plot(
    data = dataframe_of_pies, aes(x = x, y = y, label = plot),
    vp.width = 1/20, hjust = 0.5, vjust = 0.5, alpha = 0.5
  ) +
  geom_label_repel(data = site_less_than_90lon, aes(x = lon, y = lat, label = site), size = 2.5,min.segment.length = 0.01) +
  geom_label(data = site_more_than_90lon, aes(x = lon, y = lat, label = site), size = 2.5) +
  coord_map() +
  theme_cowplot()
```

You can also access a high resolution shoreline dataset for Lake Superior directly from the source() command as `lake_superior_shoreline`:

```{r, fig.width = 8, fig.height = 4}
shore <- lake_superior_shoreline

wide_view <- ggplot(shore) +
    geom_point(aes(y = lat, x = lon), size = 0.01) +
    coord_map() +
    theme_minimal()

zoom_view <- ggplot(filter(shore, lat < 47.2, lat > 46.6, lon < -90)) +
    geom_point(aes(y = lat, x = lon), size = 0.01) +
    coord_map() +
    theme_minimal()
    
plot_grid(wide_view, zoom_view, nrow = 1, rel_widths = c(1,2))
```


## {-}

## further reading {-}

- [State and County Maps in R with ggplot2](https://datavizpyr.com/how-to-make-us-state-and-county-level-maps-in-r/). A step-by-step tutorial showing how to create US state and county level choropleth maps using ggplot2 and the `map_data()` function.

- [Drawing beautiful maps programmatically with R, sf and ggplot2](https://r-spatial.org/r/2018/10/25/ggplot2-sf.html). A tutorial from the r-spatial project demonstrating how to use the `sf` package with ggplot2's `geom_sf()` to create publication-quality maps from spatial data.

- [ggtern: A package for the creation of ternary diagrams](https://www.jstatsoft.org/article/view/v087c03). The Journal of Statistical Software article introducing ggtern, which extends ggplot2 to support ternary diagrams for visualizing three-component compositional data.

<!-- ## exercises {-} -->

<!-- Using the `hawaii_aquifers` data set, please complete the following:

1. Choose one analyte and filter the data so only the rows for that analyte are shown.

2. Choose two of the aquifers. Are the mean abundances for your chosen analyte different in these two aquifers? Don't forget to test your data for normality and homogeneity of variance before selecting a statistical test. Use a plot to illustrate whether the means are similar or different.

3. Choose a second analyte, different from the first one you chose. Considering all the aquifers in the dataset, do any of them have the same abundance of this analyte? Again, don't forget about normality and homogeneity of variance tests. Use a plot to illustrate your answer.

4. Repeat #3 above, but switch the type of test used (i.e. use non-parametric if you used parametric for #3 and vice-versa). Compare the *p* values and *p* groups obtained by the two methods. Use a graphic to illustrate this. Why are they different? -->
