# Analyzing chromatography data with `phylochemistry`

One of the central features of `phylochemsitry` is its chromatogram processing tools.

```{r}
library(phylochemistry)
```

## Preparing the raw data and metadata

To use the `phylochemistry` chromatogram processing tools, we need two things: (i) a list of the samples we want to analyze (a monolist, metadata), and (ii) a singel file that contains their chromatograms (a .csv file, raw data). Here is how you get each of those things:

The first step is to export .CDF files from the chromatographic analysis system. These can then be converted into the ubiquitous .csv file format using the function `convertCDFstoCSVs`. This can take a while, but it only needs to be done once.

```{r}
phylochem_analysis_directory_path = "/Users/lucasbusta/Desktop/phylochemistry_example_data/"

GCMS_data_directory_contents <- dir(paste0(phylochem_analysis_directory_path, "raw_data/GCMS_data/"))
head(GCMS_data_directory_contents)
CDFs <- dir(paste0(phylochem_analysis_directory_path, "raw_data/GCMS_data/"))[grep(".CDF$", GCMS_data_directory_contents)]
head(CDFs)
paths_to_cdfs <- paste0(phylochem_analysis_directory_path, "raw_data/GCMS_data/", CDFs)
head(paths_to_cdfs)

phylochemistry::convertCDFstoCSVs(paths_to_cdfs = paths_to_cdfs, min_mz = 50, max_mz = 800, min_rt = 300, max_rt = 1000)
```

Once this is done, we need to create a monolist of the cdf_csvs. If you are keeping a good digital lab notebook, then you should have a list of all your samples. So, you can just modify this list. The monolist can contain any and all column you want, but it needs to include the following columns: "rt_offset", "baseline_window", and "path_to_cdf_csv". "rt_offset" should initially be 0 for all samples. "baseline_window" should initially be 100 for all samples. "path_to_cdf_csv" should be a path stating where the cdf_csv file for that sample is. Here is an example:

```{r}
my_analysis <- list()
my_analysis$GCMS_samples$sample_data <- readMonolist(monolist_in_path = "/Users/lucasbusta/Desktop/phylochemistry_example_data/monolists/GCMS_samples.csv")
head(experiment$GCMS_samples$sample_data)

my_analysis$GCMS_samples$sample_data$rt_offset <- 0
my_analysis$GCMS_samples$sample_data$baseline_window <- 100
my_analysis$GCMS_samples$sample_data$path_to_cdf_csv <- NA
my_analysis$GCMS_samples$sample_data$path_to_cdf_csv <- paste0(paths_to_cdfs[match(experiment$GCMS_samples$sample_data$sample_name, gsub(".CDF", "", gsub(".*raw_data/GCMS_data/", "", paths_to_cdfs)))], ".csv")
head(my_analysis$GCMS_samples$sample_data)

writeMonolist(monolist = my_analysis$GCMS_samples$sample_data, monolist_out_path = paste0(phylochem_analysis_directory_path, "monolists/GCMS_samples_updated.csv"))
```

We can now extract total ion chromatograms from the .csv files we just created. Since this can also take a while if there are lots of files, it's best to write these chromatograms to a new .csv so that we only have to do this extraction once.

```{r}
chromatograms <- phylochemistry::extractChromatogramsFromCSVs(paste0(paths_to_cdfs, ".csv"))

write.table(
  chromatograms,
  file = paste0(phylochem_analysis_directory_path, "raw_data/GCMS_data/chromatograms.csv"),
  row.names = FALSE,
  sep = ","
)
```

## Using the `phylochemistry` integrationApp

Once the raw data has been processed, we should have two things: (i) a list of the samples we want to process (a monolist), (ii) a .csv file that contains the chromatograms that correspond to those samples, as created by the extractChromatogramsFromCSVs() function. We are now ready to look at and analyze the chromatograms with the `phylochemistry` integration app. Let's load the app with the `integrationApp` function.

```{r}
# phylochemistry::integrationApp(
#   chromatograms = read.table(file = paste0(phylochem_analysis_directory_path, "raw_data/GCMS_data/chromatograms.csv"), sep = ",", header = TRUE),
#   x_axis_start = 300,
#   x_axis_end = 800,
#   samples_monolist_path = paste0(phylochem_analysis_directory_path, "monolists/GCMS_samples_updated.csv"),
#   create_new_samples_monolist = FALSE,
#   samples_monolist_subset = c(1:9),
#   peaks_monolist_path = paste0(phylochem_analysis_directory_path, "monolists/peak_list.csv"),
#   create_new_peak_monolist = TRUE,
#   zoom_and_scroll_rate = 50
# )
```

Based on this initial look, we can see that there is a retention time issue with samples 10_10_1 and 10_10_2. We can fix this using the the rt_offset line in the samples_monolist:

```{r}
experiment$GCMS_samples$sample_data$rt_offset[experiment$GCMS_samples$sample_data$sample_name %in% c("10_10_1", "10_10_2")] <- -12
writeMonolist(monolist = experiment$GCMS_samples$sample_data, monolist_out_path = paste0(phylochem_analysis_directory_path, "monolists/GCMS_samples_updated.csv"))
```

Now let's reload the app:

```{r}
# phylochemistry::integrationApp(
#   chromatograms = read.table(file = paste0(phylochem_analysis_directory_path, "raw_data/GCMS_data/chromatograms.csv"), sep = ",", header = TRUE),
#   x_axis_start = 300,
#   x_axis_end = 800,
#   samples_monolist_path = paste0(phylochem_analysis_directory_path, "monolists/GCMS_samples_updated.csv"),
#   create_new_samples_monolist = FALSE,
#   samples_monolist_subset = c(1:9),
#   peaks_monolist_path = paste0(phylochem_analysis_directory_path, "monolists/peak_list_example.csv"),
#   create_new_peak_monolist = FALSE,
#   zoom_and_scroll_rate = 50
# )
```

Once we've added those peaks, we can see the peak list by importing the monolist:
```{r}
experiment$GCMS_samples$peaklist <- readMonolist(monolist_in_path = paste0(phylochem_analysis_directory_path, "monolists/peak_list_example.csv"))
head(experiment$GCMS_samples$peaklist)
```

Looks good! Let's merge the peaklist and the some of the `experiment$GCMS_samples$sample_data` info:
```{r}
experiment$GCMS_samples$peaklist$sample_name <- experiment$GCMS_samples$sample_data$sample_name[match(experiment$GCMS_samples$peaklist$path_to_cdf_csv, experiment$GCMS_samples$sample_data$path_to_cdf_csv)]
str(experiment$GCMS_samples$peaklist, strict.width = "cut")
```

Let's normalize to the first peak, which is the internal standard:
```{r}
standard_areas <- plyr::ddply(experiment$GCMS_samples$peaklist, .(path_to_cdf_csv), summarize, standard_peak_area = area[peak_number_within_sample == 1])
experiment$GCMS_samples$peaklist$standard_peak_area <- standard_areas$standard_peak_area[match(experiment$GCMS_samples$peaklist$path_to_cdf_csv, standard_areas$path_to_cdf_csv)]
str(experiment$GCMS_samples$peaklist, strict.width = "cut")
```


Now we can do some plotting:
```{r}
experiment$GCMS_samples$peaklist$sample_type <- substr(experiment$GCMS_samples$peaklist$sample_name, 0, 5)
experiment$GCMS_samples$peaklist$analyte_concentration <- as.numeric(as.character(substr(experiment$GCMS_samples$peaklist$sample_type, 4, 5)))
experiment$GCMS_samples$peaklist$peak_number_within_sample <- factor(experiment$GCMS_samples$peaklist$peak_number_within_sample, levels = unique(experiment$GCMS_samples$peaklist$peak_number_within_sample))

library(ggplot2)
ggplot() +
  geom_point(
    data = experiment$GCMS_samples$peaklist, 
    aes(x = analyte_concentration, y = area/standard_peak_area, fill = peak_number_within_sample),
    shape = 21, size = 6
  ) +
  geom_smooth(
    data = experiment$GCMS_samples$peaklist, 
    aes(x = analyte_concentration, y = area/standard_peak_area),
    method = "lm"
  ) +
  facet_grid(.~peak_number_within_sample) +
  theme_classic()
```